---
title: "Stat184_CourseProject"
editor: source
author: Zachary Smith
date: 30 June 2025
date-format: DD MMMM YYYY
toc: true
format: pdf
---

```{r,echo=F}
library(rvest)
library(tidyr)
library(dplyr)
library(mosaic)
library(ggplot2)
library(knitr)
```

\newpage

## Introduction

In this project, I will explore the relationship between population, GDP (Gross Domestic Product), abd Olympic success across countries. I will use the following question to guide my project:

1.  What is the population distribtion across countries?
2.  How do the top five most populatied countries compare in terms of GDP?
3.  Is there a correlation between a countries population and the number of gold medals won in the Olympics?

## Background Information

I will be using three separate data sets to answer the questions. I wanted to find 3 datasets that related to one another and can find meaningful data points.

## Data Summary

My Primary data set contains a dataset with a column of country names and a second column with their respective population counts.

For my secondary data sets, I found a data set with a column of countries and their respective GDP and for the other data set, it has column of country names as well as the amount of gold/silver/bronze models they have each in their own column.

## What is the population distribution across countries?

To demonstrate the population distribution across countries, I chose to do the top 10 countries in population so the data wouldn't be cluttered.

```{r,echo=F,eval=TRUE, fig.cap= "Top 10 Most Populated Countries Graph"}
URL_pop <- "https://en.wikipedia.org/wiki/List_of_countries_and_dependencies_by_population"
pop_list <- URL_pop %>%
  read_html() %>%
  html_nodes(css = "table") %>%
  html_table(fill = TRUE)

popData <- pop_list[[1]] %>%
  rename(
    Country = "Location"
  ) %>%
  mutate(Population = as.numeric(gsub(",", "", Population))) 

clean_popData <- popData %>%
  filter(!grepl("World", Country, ignore.case = TRUE)) %>%
  arrange(desc(Population)) %>%
  head(10)

ggplot(data = clean_popData, aes(x = Country, y = Population)) + geom_bar(stat = "identity")  + aes(colour = Country) + theme(legend.position = "right") + labs(title = "Top 10 Most Populated Countries Graph")
```

As you can see by Figure 1, China and India are close in population but have more than triple the people of the third most populated country. The rest of the countries in Figure 1 are relatively close aside from Mexico and Russia being the lowest by roughly 100 million people compared to the third lowest which would be Bangladesh.

## How do the top 5 most populated countries compare in GDP?

The GDP of a country is essentially the health of the countries economy, so the higher the GDP the wealthier the country is. I'm using the GDP data set and the population data set to find if there's any correlation between GDP and population. The Table shows which countries were chosen based on population.

Here's a chart and table of the top 10 most populated countries:

```{r,echo=F,eval=TRUE,fig.cap= "GDP of Top 10 Most Populated Countries Chart"}
URL_gdp <- "https://en.wikipedia.org/wiki/List_of_countries_by_GDP_(nominal)"
gdp_list <- URL_gdp %>%
  read_html() %>%
  html_nodes(css = "table") %>%
  html_table(fill = TRUE)
clean_gdpData <- gdp_list[[3]] %>%
  select(1, 2) %>%
  rename(
    Country = 1,
    GDP = 2
  ) %>%
  filter(!Country %in% c("Country/Territory", "World")) %>%
  mutate(GDP = as.numeric(gsub(",", "", GDP)))
  
pop_gdp_combined <- inner_join(clean_popData, clean_gdpData, by = "Country") %>%
  select(Country, GDP, Population) %>%
  head(10)

ggplot(pop_gdp_combined, aes(x = Country, y = GDP)) + 
  geom_bar(stat = "identity") + labs(title = "GDP of Top 10 Most Populated Countries Chart")
```

The Table represents adds the population in addition to the GDP.
```{r,echo=F,eval=TRUE}
kable(pop_gdp_combined, caption = "GDP of Top 5 Most Populated Countries Table")
```
Based on both Figure 2 and Table 1, we can see that China and India being the highest in population by more than triple don't have the the largest GDP, but do have the second and third. The largest belonging to the United States.

## Is there a correlation between population and the amount of gold medals won?
To find a correlation between population and the amount of gold medals. I needed to join the population and medal data set together. Like the other figures, I used a bar chart, since it was the easiest visual to read in terms of my data. The Table shows which countries were chosen based on population.
```{r,echo=F,eval=TRUE, fig.cap= "Top 10 Most Populated Countries Gold Medal Count Chart"}
URL_medal <- "https://en.wikipedia.org/wiki/All-time_Olympic_Games_medal_table"
medal_list <- URL_medal %>%
  read_html() %>%
  html_nodes(css = "table") %>%
  html_table(fill = TRUE)

medalData <- medal_list[[2]] %>%
  select(1, 2) %>%
  rename(
    Country = 1,
    "Gold Medals" = 2
  ) %>%
  mutate(
    Country = gsub("\\[.*?\\]", "", Country), 
    `Gold Medals` = as.numeric(gsub("[^0-9]", "", `Gold Medals`))
  ) %>%
  arrange(desc(`Gold Medals`)) %>%
  head(10)

pop_medal_combined <- inner_join(medalData, popData, by = "Country") %>%
  select(Country, "Gold Medals", Population) %>%
  arrange(desc(`Gold Medals`)) 


ggplot(pop_medal_combined, aes(x = Country, y = `Gold Medals`)) + 
  geom_bar(stat = "identity") + labs(title = "Top 10 Countries with the most Gold Medals Chart")
```

The Table shows the population and amount of gold medals for each of the top ten countries with the most gold medals.
```{r,echo=F,eval=TRUE}
kable(pop_medal_combined, caption = "Top 10 Countries with the most Gold Medals Table")
```
Based on Figure 3 and Table 2, You can see how neither of the top 2 highest countries from Figure 1 appear in the list of the top 10 countries of the most gold medals. In fact, only one country, the United States from the top 5 most populated countries show in the chart. In fact, countries like Austria and Denmark which have the lowest population in Table 2 and Figure 3 are ten times lower in population than India and China. This effectively shows that there is no correlation between population and amount of gold medals.

## Conclusion
Working with data was difficult and tidying the data was the hardest task. Showing how to visualize the data sets were also a challenge. After a lot of trial and error, I was able to conclude the research questions and accurately create data visuals that showed that China and India have the highest population by a large margin. United States had the largest GDP by a large margin and didn't even have the largest population count. As for the correlation between population size and gold medals, there is none. 

## Work Cited

Countries and Population dataset: <https://en.wikipedia.org/wiki/List_of_countries_and_dependencies_by_population>

Countries and GDP dataset: <https://en.wikipedia.org/wiki/List_of_countries_by_GDP_(nominal)>

Countries and Gold Medals dataset: <https://en.wikipedia.org/wiki/All-time_Olympic_Games_medal_table>

\newpage

## Code Appendix

```{r codeAppend, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
# Population data
URL_pop <- "https://en.wikipedia.org/wiki/List_of_countries_and_dependencies_by_population"
pop_list <- URL_pop %>%
  read_html() %>%
  html_nodes(css = "table") %>%
  html_table(fill = TRUE)

popData <- pop_list[[1]] %>%
  rename(
    Country = "Location"
  ) %>%
  mutate(Population = as.numeric(gsub(",", "", Population))) 

clean_popData <- popData %>%
  filter(!grepl("World", Country, ignore.case = TRUE)) %>%
  arrange(desc(Population)) %>%
  head(10)
  
ggplot(data = clean_popData, aes(x = Country, y = Population)) + geom_point(size = 4)  + aes(colour = Country) + theme(legend.position = "right") + labs(title = "Top 10 Most Populated Countries Graph")

# GDP data
URL_gdp <- "https://en.wikipedia.org/wiki/List_of_countries_by_GDP_(nominal)"
gdp_list <- URL_gdp %>%
  read_html() %>%
  html_nodes(css = "table") %>%
  html_table(fill = TRUE)
clean_gdpData <- gdp_list[[3]] %>%
  select(1, 2) %>%
  rename(
    Country = 1,
    GDP = 2
  ) %>%
  filter(!Country %in% c("Country/Territory", "World")) %>%
  mutate(GDP = as.numeric(gsub(",", "", GDP)))
  
pop_gdp_combined <- inner_join(clean_popData, clean_gdpData, by = "Country") %>%
  select(Country, GDP, Population) %>%
  head(5)

kable(pop_gdp_combined, caption = "GDP of Top 10 Most Populated Countries Table")


ggplot(pop_gdp_combined, aes(x = Country, y = GDP)) + 
  geom_bar(stat = "identity") + labs(title = "GDP of Top 10 Most Populated Countries Chart")

# Olympic medals data
# Olympic medals data
URL_medal <- "https://en.wikipedia.org/wiki/All-time_Olympic_Games_medal_table"
medal_list <- URL_medal %>%
  read_html() %>%
  html_nodes(css = "table") %>%
  html_table(fill = TRUE)

medalData <- medal_list[[2]] %>%
  select(1, 2) %>%
  rename(
    Country = 1,
    "Gold Medals" = 2
  ) %>%
  mutate(
    Country = gsub("\\[.*?\\]", "", Country), 
    `Gold Medals` = as.numeric(gsub("[^0-9]", "", `Gold Medals`))
  ) %>%
  arrange(desc(`Gold Medals`)) %>%
  head(10)

pop_medal_combined <- inner_join(medalData, popData, by = "Country") %>%
  select(Country, "Gold Medals", Population) %>%
  arrange(desc(`Gold Medals`)) 

kable(pop_medal_combined, caption = "Top 10 Most Populated Countries Gold Medal Count Table")

ggplot(pop_medal_combined, aes(x = Country, y = `Gold Medals`)) + 
  geom_bar(stat = "identity") + labs(title = "Top 10 Most Populated Countries Gold Medal Count Chart")
```
